% Chapter 1 - Introduction.
% Main source: docs/vision.md.
% Target length: 4-6 pages.

\chapter{Introduction}
\label{ch:introduccion}

This work presents \textbf{UnivOrch}, a proof of concept for a
universal orchestrator of virtual machines aimed at teaching and
research environments. The system sits between users and the
underlying hypervisors and offers a single way to define, deploy and
operate large sets of virtual machines (VMs), regardless of whether
they run on VMware, Proxmox or a future platform. Its goal is not to
replace the hypervisors but to remove the barrier they impose on
non-expert users, in particular professors and students who only
need a small subset of what a full virtualization platform exposes.

The work is motivated by the day-to-day situation at the Computer
Engineering School (E.T.S. de Ingeniería Informática) of the
University of Málaga, where many courses depend on virtual labs and a
small team of administrators carries the full operational load.
Professors define the VMs that their subject needs and ask the
administrator for one set per student; students cannot start or stop
their own machines, and any platform change forces the administrator
to rewrite the scripts that keep everything running. The TFG starts
from two Python libraries written by the supervisor for this exact
problem, \texttt{esxobjects} and \texttt{yamlinfr}, and generalizes
them into an orchestrator that is no longer bound to VMware and that
exposes a clean public interface for other tools to build upon.

What UnivOrch contributes is best described by what it lets people do,
not by any single internal feature. Two outcomes stand out.

First, the institution stops depending on a single hypervisor. The
existing solution at the school is tied to VMware; UnivOrch puts every
hypervisor behind a common connector contract, so VMware, Proxmox and,
in the future, Hyper-V, KVM or a cloud provider can coexist and new
ones can be added without touching the core. An institution is no
longer captive to one vendor: it can run VMware in the teaching lab
and Proxmox in the research lab under the same tool, and migrate
without rewriting anything.

Second, people without technical knowledge operate virtual machines on
their own. A professor sets up a whole subject --- a folder of machines
for every enrolled student --- with a couple of commands, and the
students start, stop and reset their machines without going through the
administrator. One person manages hundreds of VMs without effort that
grows in proportion.

These outcomes rest on an architecture designed to scale and to grow.
A tree of descriptors with cascade inheritance lets the administrator
define the catalogue once and have every node below reuse it, which is
what makes hundreds of near-identical machines manageable; the same
mechanism is the natural place for a future permission model, and the
strict separation between a generic engine and a domain-specific layer
means the teaching application could be replaced by a CTF organizer or
a research-lab deployer over the same core. None of these ingredients
is individually novel --- hierarchical organizations exist in
OpenNebula and CloudStack, multi-hypervisor abstraction has been
provided by libvirt for over a decade, declarative provisioning is the
model behind Terraform. The contribution is their combination in a
lightweight, single-container tool aimed squarely at delegating VM
operation to non-technical users while staying independent of the
hypervisor.

This chapter sets the stage. \Cref{sec:motivacion} describes the
operational problem the system addresses and the context in which the
TFG is carried out. \Cref{sec:objetivos} lists the objectives this
proof of concept commits to. \Cref{sec:ajustes} discusses how the
final scope differs from the original proposal and why these
adjustments preserve the value of the work. \Cref{sec:estructura}
guides the reader through the remaining chapters and appendices.

%==================================================================
\section{Motivation}
\label{sec:motivacion}

Practical laboratory work in higher education increasingly relies on
virtualization. Reproducing a specific environment for every student
of every course is hard to do on physical hardware, while a hypervisor
can clone a base machine in seconds and discard it when the practice
is over. The technical solution is therefore well understood, but the
operational solution is not.

In the current lab at the EI UMA, the administrator is the bottleneck.
For every subject, the professor describes the VMs that their students
will need, and the administrator manually deploys one identical set of
VMs per student. The administrator also assigns IP addresses, keeps
track of which student owns which machine, intervenes when a student
loses their VM and reverts the environment after the course is over.
Each new course adds linearly to this workload. When the underlying
platform changes, in our case from a VMware version to the next, the
custom scripts that automated parts of the process need to be
rewritten from scratch.

The supervisor of this work, Dr. Guillermo Pérez Trabado, had
addressed parts of this problem with two Python libraries developed
inside the department:

\begin{itemize}
  \item \texttt{esxobjects} provides a high-level interface to the
        VMware vSphere SOAP API and hides much of its complexity from
        the rest of the code.
  \item \texttt{yamlinfr} reads a YAML description of a virtual
        infrastructure and deploys it through \texttt{esxobjects},
        with help from an external IP-Address-Management (IPAM)
        service.
\end{itemize}

These libraries cover an important part of the problem but stop
short in three ways. First, they are tied to VMware: the rest of the
school cannot use them on the Proxmox cluster used by the
infrastructure-of-systems courses. Second, they offer no abstraction
above a single deployment: there is no concept of organizing tens or
hundreds of VMs by subject, by group or by student, and no inheritance
mechanism to reuse common definitions. Third, the administrator is
still the operator: there is no service that professors can use on
their own, and certainly no view that students can use to manage their
own machines without help.

UnivOrch is built on the experience of these two libraries but
re-engineered around a different premise. The orchestrator is a long
running service, not a script; it is agnostic of the hypervisor; it
maintains its own model of the deployed infrastructure organized as a
tree with inheritance; and it exposes a public REST interface that
both human users (through a CLI today and a web application in the
future) and machines (CI/CD pipelines, GitOps tools, the future
teaching-specific layer) can call. The administrator declares the
catalogue of hypervisors and templates once; from then on, professors
and students operate through their own role-appropriate views,
without contacting the administrator for every action.

%==================================================================
\section{Objectives}
\label{sec:objetivos}

The objectives of this proof of concept are the following.

\begin{description}
  \item[\textbf{O1.} Hypervisor-agnostic orchestrator core.] \mbox{}\\
    Design and implement a service that manages a model of virtual
    machines (descriptors) and folders, organized in a tree, with no
    direct dependency on any specific hypervisor.
  \item[\textbf{O2.} Declarative model with inheritance.] \mbox{}\\
    Define a YAML-based language that allows administrators and
    professors to describe complete groups of VMs as data, with
    cascade inheritance of common properties and explicit imports for
    templates and hypervisor resources.
  \item[\textbf{O3.} Connector contract for new hypervisors.] \mbox{}\\
    Establish an abstract interface between the core and the
    hypervisor connectors so that new hypervisors can be plugged in
    without modifying the orchestrator itself.
  \item[\textbf{O4.} REST API as the public boundary.] \mbox{}\\
    Expose all orchestration operations through a REST API documented
    by an OpenAPI specification, suitable for integration with
    external tools and pipelines.
  \item[\textbf{O5.} Command-line client and single-line installation.] \mbox{}\\
    Provide a CLI client that consumes the same REST API as any other
    client, and package the daemon as a Docker image that can be
    installed with a single command, including its persistent storage.
  \item[\textbf{O6.} Engineering quality and discipline.] \mbox{}\\
    Build the proof of concept under a strict test-driven development
    process with high test coverage, static type checking and
    continuous integration, so the code base is solid enough for the
    department to keep extending after this work.
  \item[\textbf{O7.} Groundwork for the pieces beyond this PoC.] \mbox{}\\
    Leave designed (even where not fully implemented) the extension
    points for the next deliverables: real connectors against the
    teaching VMware cluster, a basic web interface and the
    teaching-specific layer-2 application.
\end{description}

A second use of these objectives appears in
\Cref{ch:conclusiones}, where each one is reviewed against what was
actually delivered.

%==================================================================
\section{Adjustments from the original proposal}
\label{sec:ajustes}

The official project proposal (\emph{anteproyecto}) listed eight
specific objectives, sketched as the natural breakdown of a complete
orchestrator. The supervisor and the author agreed from the start
that delivering all of them in 296 hours would not be realistic. The
proposal itself acknowledged this and described the work as a
\emph{proof of concept} whose purpose is to prove the design and
leave a foundation for future extensions.

\Cref{tab:adjustments} maps each objective of the proposal to its
actual status at the end of this work. The adjustments fall into
three broad categories:

\begin{itemize}
  \item \textbf{Delivered as planned.} The orchestrator engine, the
        declarative model with inheritance, the connector contract,
        the REST API, the CLI and the packaged distribution are all
        in place and ready to use.
  \item \textbf{Designed but deferred.} Several pieces have a closed
        design and the extension points are wired, but the
        implementation is left for the next iteration. The clearest
        case is the role-based access control: the model is detailed
        in the architecture documents and in DEC-011, DEC-021 and
        DEC-026, but only the \emph{superuser} role is implemented in
        code. A similar situation applies to the web interface and
        to the teaching-specific layer-2 application: their design
        is part of the architecture, but their construction is
        scheduled for the next sprints.
  \item \textbf{Blocked by external dependencies.} Real hypervisor
        connectors require access to the VMware cluster of the
        teaching laboratory. That access was not available during the
        development period, so the implementation is replaced by a
        mock connector that behaves identically from the point of
        view of the orchestrator. The connector contract was
        validated against the mock; integrating the real VMware
        connector is the immediate next task once access is granted.
\end{itemize}

\begin{table}[ht]
  \centering
  \small
  \begin{tabular}{p{0.45\linewidth} p{0.18\linewidth} p{0.30\linewidth}}
    \toprule
    \textbf{Objective from the proposal} & \textbf{Status} & \textbf{Note} \\
    \midrule
    Descriptor model with full life-cycle states &
    Delivered &
    Implemented as a two-axis state machine: descriptor state plus
    runtime state queried on demand. \\
    Linux service with central registry of descriptors &
    Delivered &
    REST daemon \texttt{univorchd} packaged as a Docker image. \\
    Hierarchical tree with cascade inheritance and permissions &
    Delivered (permissions designed) &
    Tree, inheritance and template closure implemented; RBAC designed
    but only the superuser role is wired in code. \\
    Hypervisor connectors: mock, VMware, future Proxmox &
    Partial &
    Mock connector delivered. VMware and Proxmox connectors are
    blocked by external access and scheduled as the immediate next
    task. \\
    Role-based access control with inheritance &
    Designed, not built &
    Full design closed (DEC-011, DEC-021, DEC-026). Implementation
    deferred. \\
    Teaching-specific deployment with IPAM &
    Deferred &
    Layer-2 application and IPAM integration documented and left as
    future work. \\
    REST API and web interface &
    Partially delivered &
    REST API delivered and used by the CLI. Web interface
    (NiceGUI-based) deferred to the next sprint. \\
    Coherence between descriptors and real VMs &
    Delivered (on demand) &
    Validation on load, drift detection on demand. A continuous
    reconciliation loop is documented as future work. \\
    \bottomrule
  \end{tabular}
  \caption{Mapping between the eight specific objectives of the
    project proposal and their status in this proof of concept. None
    of the items has been dropped from the design; the deferred ones
    have the necessary extension points in place.}
  \label{tab:adjustments}
\end{table}

In summary, the proof of concept covers the technical foundation end
to end and leaves the remaining pieces with a clear path to follow.
The design of the deferred items is not an afterthought: the
architecture chapter documents the extension points (the connector
abstract base class, the layer-2 boundary, the resolver of inherited
permissions, the session and IP-pool repositories) that allow each
deferred piece to be added without rewriting the core.

%==================================================================
\section{Report structure}
\label{sec:estructura}

The rest of this report is organized as follows.

\Cref{ch:conceptos} (\textbf{System concepts}) introduces the
ideas the reader needs in order to follow the rest of the report
without gaps. It describes UnivOrch from the point of view of
its three kinds of user, presents the descriptor tree and the
cascade inheritance, explains the separation between the engine
and the domain application, and walks through the YAML model
with worked examples.

\Cref{ch:estado-arte} (\textbf{State of the art}) reviews the
technical context: classic teaching virtual labs, current
multi-hypervisor abstractions, declarative configuration systems,
hierarchical access control and the specific case of container-based
versus VM-based labs. It positions UnivOrch against the closest
academic and industrial precedents and identifies the niche it
occupies.

\Cref{ch:metodologia} (\textbf{Methodology}) describes the working
method: the seven-phase plan adapted from a SCRUM-like iterative
process, the test-driven development cycle, the supporting tooling
(devcontainer, continuous integration, code quality gates) and the
use of AI-assisted development as a deliberate and traceable part of
the workflow.

\Cref{ch:desarrollo} (\textbf{Development, testing and results}) is
the technical core of the report. It presents the requirements
analysis, the architecture of the system in detail and the
implementation work organized by layer, followed by the testing
strategy and a functional demonstration end to end.

\Cref{ch:conclusiones} (\textbf{Conclusions and future work}) reviews
the objectives against the delivered system, summarizes the
contributions and limitations, and outlines the natural extensions of
the work.

Five appendices complete the report. Appendix~\ref{ap:manual-admin}
contains the administrator's manual. Appendix~\ref{ap:requirements}
transcribes the detailed requirements analysis.
Appendix~\ref{ap:decisiones} indexes the design decisions referenced
throughout the text. Appendix~\ref{ap:cli-reference} provides a CLI reference.
Finally, Appendix~\ref{ap:conclusiones-es} reproduces the conclusions
in Spanish, as required by the regulations of the EI UMA when the
report is written in English.
