% Chapter 2 - State of the art and technologies.
% Sources: docs/memoria/_notes/state-of-the-art-research.md,
% docs/technologies.md.
% Target length: 8-10 pages.

\chapter{State of the art and technologies}
\label{ch:estado-arte}

This chapter places UnivOrch in the landscape of existing systems
and ideas. We approach it from several angles: the closest academic
precedent of virtual labs for higher education, the lineage of
multi-hypervisor abstractions, the broader move from imperative to
declarative infrastructure management, the well-known field of
role-based access control with hierarchy, and a recurring discussion
that surrounds any educational lab today, namely whether containers
or VMs are the right substrate. We then describe the two libraries
that the supervisor had developed before this work and that serve as
the operational starting point. We close with an honest positioning
of UnivOrch against everything that exists, and with the concrete
technologies adopted for the implementation.

The aim throughout is to be precise about what is already available
and what is not. Several pieces of UnivOrch borrow from prior art and
the report acknowledges those debts; the contribution lives in the
specific combination, not in any individual feature.

%==================================================================
\section{Virtualization, orchestration and the teaching case}
\label{sec:virtualizacion-orquestacion}

The terms \emph{hypervisor}, \emph{hypervisor manager} and
\emph{orchestration} are often used loosely. In this work we use them
in a strict sense. A \emph{hypervisor} (VMware ESXi, Proxmox VE,
KVM, Hyper-V) creates and runs virtual machines on a single host. A
\emph{hypervisor manager} (VMware vCenter, Proxmox cluster) groups
several hypervisors of the same family under a single administrative
view, but still speaks the vocabulary of the hypervisor. An
\emph{orchestrator} sits one level above: it manages the life cycle
of VMs across one or more managers, exposes its own model of the
resources, and can speak to several heterogeneous hypervisors through
adapters.

The orchestrator can be \emph{imperative} (the user issues
\texttt{create}, \texttt{start}, \texttt{stop} commands) or
\emph{declarative} (the user describes the desired state and the
orchestrator computes the operations needed to reach it). Burgess
established the theoretical framing of this latter model in his
\emph{theory of system administration}~\cite{burgess2003theory},
introducing the notions of \emph{convergence} between desired and
actual state and \emph{configuration drift}, which form the
conceptual basis of Terraform, Ansible and Kubernetes alike. UnivOrch
adopts the declarative model with on-demand application;
continuous reconciliation is mentioned in the closing chapter as a
natural extension.

The case that justifies a dedicated orchestrator is well documented
in the academic literature on virtual labs. The most directly
comparable academic precedent is the Virtual Computing Laboratory
(VCL) of North Carolina State University, presented in IEEE
Computer~\cite{vouk2009ncsu} and described in detail
in~\cite{averitt2007vcl, vouk2008cloud}. The VCL is an open-source
cloud orchestrator built explicitly for teaching and research, used
by tens of thousands of students. It demonstrates that a single
service that lets professors define VM environments and that lets
students access their own machines is feasible. UnivOrch differs from
the VCL in three concrete aspects: it is built as a
hypervisor-agnostic engine instead of a tightly integrated stack, it
uses a declarative YAML model with cascade inheritance instead of an
imperative reservation system, and it separates the domain layer
(teaching, CTF, research) from the engine through a clean public
boundary.

%==================================================================
\section{Multi-hypervisor abstraction layers}
\label{sec:multi-hv}

Abstracting several hypervisor families behind a common interface is
not a new idea. The reference is
\texttt{libvirt}~\cite{bolte2010libvirt}, an open-source library that
exposes VMware, Hyper-V, KVM and Xen behind a single API. Its model
inspired the abstract \texttt{HypervisorConnector} interface used in
UnivOrch (\Cref{ch:desarrollo}), and the connector contract follows
the same general shape: a small set of life-cycle operations
(\texttt{clone}, \texttt{delete}, \texttt{start}, \texttt{stop},
\texttt{get\_status}) plus a way to expose hypervisor-specific
parameters.

At a higher level, several open-source orchestrators already provide
multi-hypervisor coverage. OpenNebula~\cite{opennebula-docs} supports
KVM, LXC, vSphere and Firecracker and organizes users in two-level
groups (groups and Virtual Data Centers); Apache
CloudStack~\cite{cloudstack-docs} supports KVM, vSphere, XenServer
and Hyper-V and organizes both the infrastructure (Region $>$ Zone
$>$ Pod $>$ Cluster) and the user space (Domain $>$ Account) in
trees; oVirt~\cite{ovirt-docs} focuses on KVM but exposes Data
Center $>$ Cluster $>$ Host as a hierarchy with inheritance of
machine type and CPU model from the cluster. The Distributed
Management Task Force, on the standards side, offers Open
Virtualization Format (OVF)~\cite{dmtf-ovf} for cross-hypervisor VM
portability.

The conclusion is that multi-hypervisor abstraction \emph{per se} is
not a novel contribution; what UnivOrch contributes in this area is a
much lighter integration model. OpenNebula and CloudStack are
substantial systems that assume a dedicated operations team. UnivOrch
keeps the same connector pattern but ships as a single container
intended to be installed in one command and operated by a single
person.

%==================================================================
\section{Declarative infrastructure management}
\label{sec:declarative}

The declarative model has become the standard in the
infrastructure-as-code field. Rahman, Mahdavi-Hezaveh and Williams
present a systematic mapping study~\cite{rahman2019mapping} that
identifies the main research themes in this area, and their companion
work on defective IaC scripts~\cite{rahman2019defective} gives
empirical grounding to the discussion of correctness. The dominant
industrial tool, Terraform, organizes infrastructure in
\emph{modules} that can be composed through inputs and outputs, but
variables do not inherit across module boundaries automatically;
inheritance has to be wired by hand.

The closest current system to what UnivOrch tries to achieve is
Crossplane~\cite{crossplane-docs}, which runs inside a Kubernetes
cluster and uses Composite Resource Definitions (XRDs) and
Compositions to abstract any cloud or platform resource behind a
simplified Claim. Crossplane resolves elegantly the separation
between a generic engine (Kubernetes plus the Crossplane controllers)
and a domain layer (the XRDs that each organization defines for its
own platform). This work cannot ignore Crossplane: the architectural
pattern is similar in spirit. UnivOrch differs in three points: it
does not require a Kubernetes cluster, which is a practical advantage
in environments with one to a few hosts; it uses a tree-structured
domain model with cascade inheritance instead of composition graphs;
and it accepts that it is much less mature and less extensible than
Crossplane.

In the specific niche of \emph{GitOps over VMs}, Red Hat has
documented an integration of KubeVirt with OpenShift
GitOps~\cite{kubevirt-gitops} that treats VM definitions as code in a
Git repository. This is a useful reference for the future-work
section of this report; UnivOrch does not implement Git triggers in
v1 but its declarative YAML interface and the public REST API leave
the door open to the same workflow.

%==================================================================
\section{Hierarchical RBAC and tenant delegation}
\label{sec:rbac}

Role-Based Access Control with role hierarchy is a classical topic,
formalized by Sandhu et al.\ in~\cite{sandhu1996rbac} and later
standardized as the NIST RBAC model~\cite{ferraiolo2001nist} and the
\mbox{ANSI INCITS}~359-2004 standard~\cite{ansi2004rbac}. The RBAC1
model directly maps to what UnivOrch implements at the design level:
a hierarchy of roles in which the assignment at a higher level
cascades to descendants, with the possibility of overriding at any
node.

Delegation of management to non-administrator users with a
simplified view is also common in industry products. OpenNebula
exposes a \emph{Cloud View} for end users in its Sunstone interface;
VMware Cloud Director offers Tenant Portals; Nutanix Prism
incorporates Self-Service blueprints. In all these systems, an
ordinary user receives a curated subset of operations corresponding
to their role, without exposure to the underlying machinery. UnivOrch
embraces this same principle and goes further in one specific
dimension: the metaphor presented to the student is intentionally
domain-specific (a \emph{desk} that contains \emph{computers}) and
hides not only the technical machinery but also the orchestrator's
own concept of a tree. This is a contribution at the level of
user-experience design, not at the level of access-control
mechanism; we present it as such in \Cref{ch:desarrollo}.

The closest pattern in the open-source space is the Hierarchical
Namespace Controller (HNC) of Kubernetes~\cite{k8s-hnc-blog}, which
introduces a tree of namespaces with automatic propagation of RBAC
roles, resource quotas and network policies. HNC is also a good
reference for the inheritance semantics chosen in UnivOrch.

%==================================================================
\section{Lexical scoping applied to infrastructure}
\label{sec:lexical}

One specific aspect of UnivOrch deserves separate discussion because
it is, to our knowledge, less common in the orchestration literature
and more familiar to programming-language theory. When a template
is imported from a parent folder into a child folder, references
inside the template (\textit{e.g.}\ the hypervisor it targets) are
resolved \emph{against the folder where the template was originally
defined}, not against the folder where it is used. This is exactly
the behaviour of a \emph{lexical closure} in a programming language:
the environment in which a function was defined accompanies the
function wherever it travels.

The notion of lexical scoping was introduced by Sussman and Steele in
the seminal Scheme paper~\cite{sussman1975scheme} and is the subject
of a clear pedagogical treatment in Chapter~3 of
SICP~\cite{abelson1996sicp}. In the configuration-language space, the
idea is alive in Jsonnet~\cite{jsonnet-lang}, which supports lexical
scoping and object inheritance with \texttt{self} and \texttt{super}
keywords, and in similar languages such as CUE and Dhall.

The novelty of UnivOrch in this respect is modest but real: this
work applies the lexical-closure pattern to a concrete model of
infrastructure resources, integrated with a hierarchical organization
of users and a multi-hypervisor adapter layer. We have not found
this combination in any production orchestrator surveyed.

%==================================================================
\section{Containers and virtual machines in educational labs}
\label{sec:vm-vs-container}

A reasonable reader may ask, in 2026, why a TFG on VM orchestration
makes sense at all when container-based stacks are dominant. The
answer comes from the specific requirements of practical work in
computer science education. Several recent works are explicit on the
matter. The KYPO cyber range~\cite{vykopal2017kypo} argues that
hands-on training in network security and operating-system
internals demands access to a privileged kernel and to arbitrary
networking, neither of which a non-privileged container can offer.
The CyTrONE framework~\cite{beuran2018cytrone} compares both
substrates and reaches the same practical conclusion: containers
suffice for many introductory scenarios, but full VMs remain
necessary for kernel-level and complex-networking exercises. The
Tele-Lab project~\cite{willems2012telelab} of HPI Potsdam is an
earlier example of a VM-based learning environment with a long
operational history.

The Labtainers framework~\cite{irvine2017labtainers}, in contrast,
demonstrates that the container approach is perfectly viable when
the lab focuses on application-level security and does not require
kernel modifications, and it offers an interesting model of
parameterized, per-student lab instances.

The conclusion is that the choice of substrate depends on the
nature of the lab. UnivOrch focuses on the VM case for two reasons:
the existing labs at the EI UMA rely on VMware ESXi and Proxmox; and
the kinds of exercises that drive the most operational pain at the
school (computer networks, operating systems, infrastructure
administration) are precisely the ones that require full VMs.

%==================================================================
\section{Starting point: the supervisor's previous work}
\label{sec:punto-partida}

The TFG starts from two Python libraries developed by the supervisor,
Dr.~Guillermo Pérez Trabado, inside the Department of Computer
Architecture of the University of Málaga. They are unpublished
internal software but solve a real operational need at the school.

\texttt{esxobjects} is a Python library that wraps the VMware
vSphere SOAP API behind a friendlier object-oriented interface. It
takes care of authentication, session management and the most
common operations on VMs and templates. It is tied to VMware by
construction.

\texttt{yamlinfr} is a higher-level layer built on top of
\texttt{esxobjects}. It reads a YAML description of an infrastructure
of VMs and deploys it onto VMware. It also interacts with an
external IPAM service to allocate IP addresses. It is the closest
operational antecedent of UnivOrch: it already adopts a YAML-based
declarative model and the idea of deploying a whole infrastructure
at once.

UnivOrch generalizes this experience along three axes. First, the
hypervisor is no longer fixed: a connector abstract class allows
VMware, Proxmox and a development mock to coexist behind the same
interface. Second, the YAML model is enriched with a tree of folders,
cascade inheritance and template closures, which the previous
libraries did not provide. Third, the system becomes a long-running
service with a public REST API, accessible to multiple clients
(CLI, future web GUI, external tools), rather than a one-shot
deployment script. A qualitative comparison between
\texttt{esxobjects}, \texttt{yamlinfr} and UnivOrch closes
\Cref{ch:desarrollo}.

%==================================================================
\section{Positioning of UnivOrch}
\label{sec:positioning}

Summarizing the sections above, the original contribution of
UnivOrch is not the appearance of any individual feature in
isolation. Hierarchical organizations of resources and users are
provided by OpenNebula, CloudStack, vCloud Director, oVirt and the
Kubernetes HNC; declarative provisioning is the standard model
behind Terraform and the various flavours of GitOps; multi-hypervisor
adapters have been provided by libvirt for over a decade; lexical
scoping in configuration languages already exists in Jsonnet, CUE
and Dhall; tenant delegation with simplified user views is offered
by every major commercial cloud platform.

What UnivOrch contributes is the synthesis of these ingredients in a
lightweight, single-container service explicitly aimed at the
teaching and small-research case. Concretely, three points are
defensible as original:

\begin{enumerate}
  \item The application of a lexical-closure pattern to
        infrastructure templates inside a hierarchical tree with
        cascade inheritance by data type. We have not found this
        combination in the surveyed production systems.
  \item A strict separation between a domain-agnostic engine and a
        domain-specific layer (teaching, CTF, research), implemented
        as a thin client of a public REST API rather than as a
        controller on top of Kubernetes. The pattern itself is
        familiar from Crossplane; the contribution is to show that
        an equivalent decoupling is achievable in a much lighter
        deployment.
  \item A user-experience model in which non-technical users
        (students) operate VMs through a domain-specific metaphor
        (\emph{desks} and \emph{computers}) that hides not only the
        hypervisor but the orchestrator's own concepts. This is a
        design contribution rather than an engineering one and we
        acknowledge it as such.
\end{enumerate}

The closest single point of reference among all the systems
surveyed is the NCSU VCL~\cite{vouk2009ncsu}; the largest unfair
fight, in terms of maturity and engineering depth, is Crossplane.
UnivOrch is neither at the scale of the former nor at the maturity
of the latter; it occupies a defensible niche between them.

%==================================================================
\section{Selected technologies}
\label{sec:tecnologias-seleccionadas}

\Cref{tab:tech-stack} summarizes the technologies adopted for the
implementation. The choices are documented in detail in
\texttt{docs/technologies.md} of the project repository and in the
design-decision DEC-033; we restate the rationale here.

\begin{table}[ht]
  \centering
  \small
  \begin{tabular}{p{0.20\linewidth} p{0.30\linewidth} p{0.42\linewidth}}
    \toprule
    \textbf{Layer} & \textbf{Selected technology} & \textbf{Why} \\
    \midrule
    Language &
    Python 3.12 &
    Same language as the supervisor's libraries; native abstract
    classes, type hints and \texttt{@override}; well supported in
    a long-lived container. \\
    Web framework &
    FastAPI + Uvicorn~\cite{fastapi-docs} &
    Generates an OpenAPI specification automatically; strict typing
    integrates with Pydantic; ASGI test client is standard. \\
    Domain modelling &
    Pydantic v2~\cite{pydantic-docs} &
    Declarative model, validation at the boundary, JSON
    serialization for free, used by FastAPI internally. \\
    HTTP client (CLI) &
    httpx &
    Same API as \texttt{requests}, with async support and a strict
    test transport; replaceable through dependency injection in
    tests. \\
    Persistence &
    TinyDB~\cite{tinydb-docs} (PoC) &
    A single JSON file as the database keeps the proof of concept
    self-contained. A MongoDB-backed repository is the natural
    successor for production. \\
    CLI &
    cmd2 &
    Bash-style and REPL modes from the same command set; argparse
    integration; well-tested behaviour for tab completion and help. \\
    Quality &
    Ruff~\cite{ruff-docs}, mypy --strict,
    pytest, Hypothesis~\cite{hypothesis-docs} &
    Fast linter and formatter; strict static typing; full unit and
    integration test coverage; property-based testing on the pure
    Resolver. \\
    Tooling &
    uv~\cite{uv-docs} + Docker &
    \texttt{uv} for reproducible Python environments; multi-stage
    Docker for a minimal runtime image; image hosted in GitHub
    Container Registry. \\
    \bottomrule
  \end{tabular}
  \caption{Summary of the implementation technology stack.}
  \label{tab:tech-stack}
\end{table}

Two design alternatives are worth mentioning briefly. SQLite was
considered for persistence; it was rejected in favour of TinyDB
because the descriptor data is naturally document-shaped (deeply
nested YAML fragments), not relational, and TinyDB removes the need
for an ORM in the proof of concept. Click was considered as a CLI
framework; cmd2 was preferred because it offers an interactive REPL
in addition to one-shot commands without extra code, which matches
the experience the end user is expected to have.

The implementation is presented in detail in \Cref{ch:desarrollo}.
