% Chapter 5 - Conclusions and future work.
% Sources: claude/ideas-pendientes.md and the journal closing entries.
% Target length: 3-5 pages.
%
% Note: as the report is written in English, UMA EI regulations
% require this chapter to also appear in Spanish. The Spanish version
% is included as Appendix E (E-conclusiones-castellano.tex).

\chapter{Conclusions and future work}
\label{ch:conclusiones}

This chapter closes the report. It reviews the proof of concept
against the objectives set in \Cref{ch:introduccion}, summarizes the
main contributions and the known limitations, and outlines the
natural extensions of the work beyond this TFG. The chapter is short
on purpose: most of the underlying material has already been
defended throughout the report, so the conclusions can stay tight.

%==================================================================
\section{Conclusions}
\label{sec:conclusiones}

\subsection{Fulfilment of objectives}

The seven objectives declared in \Cref{sec:objetivos} are reviewed
here against what was actually delivered. \Cref{tab:objectives-review}
summarizes the assessment; the following paragraphs add the
qualifications that the table cannot capture.

\begin{table}[ht]
  \centering
  \small
  \begin{tabular}{p{0.06\linewidth} p{0.50\linewidth} p{0.14\linewidth} p{0.18\linewidth}}
    \toprule
    \textbf{ID} & \textbf{Objective} & \textbf{Status} & \textbf{Reference} \\
    \midrule
    O1 & Hypervisor-agnostic orchestrator core           & Delivered &
       \Cref{sec:diseno,sec:implementacion} \\
    O2 & Declarative model with inheritance              & Delivered &
       \Cref{subsec:domain-model,subsec:load-flow} \\
    O3 & Connector contract for new hypervisors          & Delivered &
       \Cref{subsec:patterns} \\
    O4 & REST API as the public boundary                 & Delivered &
       \Cref{subsec:rest-boundary} \\
    O5 & CLI client and single-line installation         & Delivered &
       \Cref{sec:implementacion} \\
    O6 & Engineering quality and discipline              & Delivered &
       \Cref{sec:pruebas,ch:metodologia} \\
    O7 & Groundwork for the pieces beyond this PoC       & Partial   &
       \Cref{sec:futuro} \\
    \bottomrule
  \end{tabular}
  \caption{Status of the seven objectives at the close of the proof
    of concept.}
  \label{tab:objectives-review}
\end{table}

Objectives O1 through O6 are met by the delivered system: the
engine runs, the declarative model with cascade inheritance and
template closure works end-to-end, the connector contract is
defined and exercised by the mock implementation, the REST API
exposes every public operation, the CLI is a pure HTTP client of
the same API, and the engineering practices (test-driven
development, strict static typing, continuous integration, packaged
distribution) hold across the whole project.

Objective O7 is marked \emph{partial}: every extension point is
designed and most of them are wired in code, but the next
deliverables that depend on them (real connectors, web GUI, teaching
application) are not implemented in this TFG. The architecture
chapter and the design-decisions log document the extension points
precisely enough that the work to come can proceed without
revisiting the engine.

\subsection{Main contributions}

The contributions of this work are honest about the state of the
art (\Cref{ch:estado-arte}). The PoC does not introduce
fundamentally new techniques; it integrates known techniques in a
specific combination that, to our knowledge, no other production
system provides:

\begin{description}
  \item[A descriptor tree with lexical-closure templates.]
        The combination of cascade inheritance by data type and
        lexical-closure resolution of imported templates is the
        single most distinctive contribution. The pattern, familiar
        from programming languages, is applied here to a concrete
        infrastructure model.
  \item[A clean separation between engine and domain layer.]
        The two-layer architecture allows the same engine to be
        used for teaching, CTF competitions, research labs or any
        other use case that can be expressed as a tree of VM
        descriptors. The pattern is similar in spirit to
        Crossplane on Kubernetes; the contribution is to show that
        an equivalent decoupling is achievable in a lightweight
        single-container deployment.
  \item[A domain-specific user model.]
        The end user (student) operates VMs through a
        domain-specific metaphor (\emph{desks} and
        \emph{computers}) that hides not only the hypervisor but the
        orchestrator's own concept of a tree. This is a
        user-experience contribution rather than an engineering one,
        and the report acknowledges it as such.
  \item[A proof of concept with full engineering discipline.]
        The delivered code base, despite being a PoC, ships with
        244 automated tests, branch coverage above 95\% on the
        domain logic, strict static typing, continuous integration
        gates, a published image in the GitHub Container Registry
        and a tutorial for the operator. The discipline is itself a
        deliverable: the future iterations of the project will
        inherit a code base that is genuinely maintainable.
\end{description}

\subsection{Known limitations}

The limitations of the proof of concept are listed openly so that
the assessment of the work is not based on optimistic assumptions.

\begin{itemize}
  \item \textbf{No real hypervisor connector.} The mock connector
        proves the contract, but a deployment against a real VMware
        or Proxmox host is not implemented. The bottleneck is
        external: VPN access to the teaching VMware cluster was
        not available during the development period. The VMware
        connector is the immediate next task once access is
        granted.
  \item \textbf{No graphical interface.} The CLI is the only client
        delivered. The web GUI was deferred to the next sprint
        because the REST boundary, the CLI and the engine had
        higher priority for the PoC.
  \item \textbf{Teaching application not implemented.} The
        layer-2 teaching application is fully designed but the
        implementation is left for a later sprint. The current
        operator workflow goes through the generic CLI; an
        operational deployment of a real course is therefore not
        as smooth as the design intends.
  \item \textbf{Authentication not implemented.} The RBAC model is
        fully designed (DEC-011, DEC-021, DEC-026) and only the
        superuser role is realistic in v1. The service assumes it
        runs in a trusted network. Authentication is part of the
        future-work list.
  \item \textbf{Best-effort persistence.} TinyDB does not support
        multi-document transactions. The cost of a crash in the
        middle of a multi-document write is accepted explicitly in
        DEC-030; the replacement by MongoDB lifts this limitation.
  \item \textbf{Single-host service.} The PoC assumes a single
        host running the daemon. Active/passive high availability
        is designed but not implemented.
\end{itemize}

%==================================================================
\section{Future work}
\label{sec:futuro}

The following extensions are the natural continuation of the work,
ordered from the most immediate to the most ambitious. Most of them
were anticipated in the design and have their extension points in
place; they have been documented in
\texttt{claude/ideas-pendientes.md} throughout the development.

\subsection{Next sprints (immediate)}

\textbf{Real hypervisor connectors.} The VMware connector is the
highest priority, both because it unblocks an operational deployment
against the existing teaching lab and because it validates the
abstract connector contract against the messiest of the two target
APIs (vSphere SOAP). The Proxmox connector follows the same plan
once VMware is in.

\textbf{Basic web interface.} A NiceGUI-based interface that uses
the REST API as a thin client. The first version focuses on
read-only navigation of the tree, descriptor details, job log and
state at a glance. Write operations (create, edit, deploy) are added
in a second iteration.

\textbf{Teaching application (layer 2).} A single command takes a
list of students and a base template and produces a complete
subject deployment: one folder per student, one descriptor per
student, the right cascade configuration. A second command recycles
the whole subject at the end of the term.

\textbf{Authentication and RBAC.} Implement the role-based access
control already designed. Users in a YAML file in v1; an LDAP /
Active Directory backend behind the \texttt{UserRepository}
abstraction in v2. Session tokens persisted in the database, as
designed in DEC-031.

\subsection{Beyond the immediate roadmap}

\textbf{Richer web interface.} Operational dashboards, advanced
editing of the tree, alternative views from the hypervisor side.

\textbf{IPAM integration.} A real IPAM (\emph{e.g.}\ phpIPAM,
NetBox) replaces the internal IP-pool model. The
\texttt{IPPoolRepository} is the seam.

\textbf{Snapshot management.} Snapshots through the orchestrator
with inheritable policies (retention, automatic-snapshot triggers).

\textbf{Datastores modelled as inheritable resources.} The current
proof of concept hardcodes the default datastore of each
hypervisor. A first-class datastore model with alias-based
references, similar to the existing template model, brings storage
policy into the declarative layer.

\textbf{Additional connectors.} Hyper-V, KVM, public-cloud
providers (AWS EC2, Azure VM). Each is a separate connector
implementing the same abstract base class.

\textbf{High availability.} Active/standby deployment of the
daemon, with MongoDB replacing TinyDB to provide multi-document
transactions and replication. The session and job models were
designed with this future in mind.

\textbf{GitOps over VMs.} The declarative YAML model lends itself
to a Git-backed workflow: the tree of definitions lives in a Git
repository, a push triggers \texttt{load}, and the orchestrator
becomes the convergence engine. Equivalent to what ArgoCD or Flux
do for Kubernetes, but for VM infrastructure.

\textbf{Structured logging to syslog.} The current logging is
ad-hoc to standard output. A structured pipeline through the
Python \texttt{logging} module to syslog is designed in DEC-023
and pending implementation.

\textbf{Safety policies on destructive operations.}
\texttt{undeploy} currently succeeds on a running VM by stopping
it first. A \texttt{--force} flag and a refuse-by-default policy
on running VMs is part of the future safety work documented in
\texttt{claude/ideas-pendientes.md}.

%==================================================================

The proof of concept proves the model. The path to a production
system goes through the items above, in roughly the order listed.
The engine is ready to receive them.
