% Appendix A - Administrator's manual.
% Sources: docs/tutorial-profesor.md, docs/environment.md, the
% install.sh script and the README of the repository. Update here when
% the operational story changes.

\umaappendix{Administrator's manual}
\label{ap:manual-admin}

This appendix is the operational manual for an administrator who
installs and runs UnivOrch in an institutional environment. It does
not repeat the conceptual material of \Cref{ch:conceptos}; the
reader is assumed to have followed the rest of the report. The
manual covers requirements, installation, day-to-day operation,
configuration of real hypervisors (designed but not implemented in
the proof of concept), monitoring and maintenance.

The conventions are: a sequence of commands separated by line
breaks is run in order; comments after \texttt{\#} are not part of
the command; placeholders in angle brackets
(\texttt{<server>}) are to be replaced.

\section{Requirements}

The host that runs UnivOrch needs:

\begin{itemize}
  \item Linux (any distribution with current packages).
  \item Docker Engine version 24 or newer.
  \item Docker Compose, version~2 (\texttt{docker compose}) or
        version~1 (\texttt{docker-compose}); the installer detects
        both.
  \item Outbound HTTPS to \texttt{ghcr.io} to pull the public
        image. If the host is in a closed network, the image can
        be mirrored to an internal registry; the configuration
        change is the registry name in \texttt{docker-compose.yml}.
\end{itemize}

The clients (CLI on a workstation, web browser on a laptop) need
only network access to the daemon's port (8080 by default).

UnivOrch does not require root: a user in the \texttt{docker}
group is enough.

\section{Quick installation}

The published installer takes care of everything for a single-host
deployment. It downloads the
\texttt{docker-compose.yml}, picks a free TCP port if 8080~is
taken, pulls the image, creates the persistent volume and starts the
container.

\begin{verbatim}
curl -sSL https://raw.githubusercontent.com/clamaveruma/univorch/main/install.sh | bash
\end{verbatim}

A health check confirms the install:

\begin{verbatim}
curl -s http://localhost:8080/api/v1/health
# {"status":"ok"}
\end{verbatim}

For a closer look at what the installer does before running it,
substitute \texttt{bash} with \texttt{less} or \texttt{cat}:

\begin{verbatim}
curl -sSL https://raw.githubusercontent.com/clamaveruma/univorch/main/install.sh | less
\end{verbatim}

\section{Production installation}

For production the recommended path is to clone the repository,
edit \texttt{docker-compose.yml} to the local conventions and run
\texttt{docker compose up -d}.

\begin{verbatim}
git clone https://github.com/clamaveruma/univorch.git
cd univorch
$EDITOR docker-compose.yml   # adjust port, volume name, image tag
docker compose up -d
docker compose logs -f       # follow the daemon's log
\end{verbatim}

The file \texttt{docker-compose.yml} exposes three pieces of
configuration:

\begin{description}
  \item[Image tag] -- pin the deployment to a specific version
        instead of \texttt{:latest}. Example:
        \texttt{ghcr.io/clamaveruma/univorch:v0.2.0}. The tags
        follow semantic versioning; \texttt{v0.x} indicates the
        API surface is not yet stable.
  \item[Port mapping] -- the public port the daemon listens on. The
        default is 8080~on the host mapped to 8080~in the
        container. Change the host side if the port is busy or if
        a reverse proxy fronts the daemon.
  \item[Named volume] -- the Docker named volume that contains the
        TinyDB JSON file. The default name is
        \texttt{univorch\_data}; keep the default unless several
        UnivOrch instances coexist on the same host.
\end{description}

The image does \emph{not} terminate TLS. In a production
deployment, place the daemon behind a reverse proxy (nginx,
Caddy, Traefik, the institutional load balancer) that handles
HTTPS, forwards the request unchanged and trusts the daemon on
HTTP. The daemon binds to all interfaces inside the container by
default; restrict the binding through the proxy, not through
UnivOrch.

\section{Real hypervisor configuration (design)}

The PoC ships only the mock connector. The real connectors
(VMware vSphere and Proxmox VE) are designed but not implemented
in v1. The configuration of a real hypervisor below describes the
intended user interface; it will appear once the connectors are
in code.

\paragraph{VMware ESXi / vSphere.}
The administrator declares the hypervisor in the top
administrative folder of the tree:

\begin{verbatim}
folders:
  admin:
    define hypervisors:
      esx1:
        type: vmware
        address: vcenter.dept.uma.es
        username: orchestrator@vsphere.local
        password: <stored encrypted>
        datacenter: TeachingLab
        datastores:
          fast:   { id: SSD-Pool-1 }
          slow:   { id: HDD-Pool-2 }
\end{verbatim}

The base VMs available for cloning live in the same vSphere
datacenter and have a snapshot ready for linked clones. The
administrator prepares the base VMs once and refers to them by
name from templates.

\paragraph{Proxmox VE.}
Proxmox uses an API token rather than a password:

\begin{verbatim}
folders:
  admin:
    define hypervisors:
      pve1:
        type: proxmox
        address: pve.dept.uma.es
        token_id: orchestrator@pve!api
        token_secret: <stored encrypted>
        node: pve-node-01
        datastores:
          local: { id: local-lvm }
\end{verbatim}

The base VM is a template VM created in Proxmox; cloning is
delegated to the standard Proxmox API.

\paragraph{Mock.}
The mock connector requires no configuration. The administrator
declares it the same way:

\begin{verbatim}
folders:
  admin:
    define hypervisors:
      mock01: { type: mock }
\end{verbatim}

\section{Loading definitions}

Definitions are written in YAML and loaded into a folder of the
tree with the \texttt{load} command. The destination can be any
folder, not only the root; the YAML is relative to the
destination (\Cref{sec:declarativo}).

The administrator typically keeps the catalogue of definitions in
a Git repository. A simple workflow is:

\begin{verbatim}
git clone git@servers.dept.uma.es:univorch-defs.git
cd univorch-defs
univorch load catalogue/hypervisors.yml /admin/
univorch load courses/networks-2026.yml /lab/
\end{verbatim}

Two YAML files describe two complementary pieces of the tree: a
catalogue of hypervisors and templates at \texttt{/admin/}, and a
subject's worth of student descriptors at \texttt{/lab/}. The
order of \texttt{load} calls matters when a later file references
resources declared by an earlier one.

\texttt{load} is idempotent. Running the same file twice does not
duplicate folders or descriptors; only new items are added.
Properties of existing items are updated to match the file.

\section{Day-to-day operation}

The administrator interacts with the daemon through the CLI or
the web GUI. The most frequent operations are listed below;
\Cref{ap:cli-reference} gives the full reference.

\paragraph{Inspecting the tree.}
The CLI is the fastest way to scan the current state. From inside
the container:

\begin{verbatim}
docker exec -it univorch univorch
univorch /> tree /
univorch /> status /lab/networks/student01
univorch /> inspect /lab/networks/student01
\end{verbatim}

The web GUI at \texttt{http://<server>:8080/} gives the same
information graphically, with the summary counters at the top of
the page.

\paragraph{Deploying and undeploying.}
A descriptor in \texttt{provisioned} becomes a real VM with
\texttt{deploy}; \texttt{undeploy} destroys the VM and returns the
descriptor to \texttt{provisioned}.

\begin{verbatim}
univorch /> deploy /lab/networks/student01
univorch /> start /lab/networks/student01
univorch /> stop /lab/networks/student01
univorch /> undeploy /lab/networks/student01
\end{verbatim}

For batch operations across a whole folder, scripting is the
practical option in v1; the web GUI will offer the same operations
on a click in the next sprint.

\paragraph{Recovering a broken descriptor.}
A descriptor stuck in \texttt{broken} reports the cause in its
Job history. Read the history, fix the underlying problem
(restore credentials, free disk space, recover network
connectivity to the hypervisor) and force-undeploy the descriptor
to bring it back to \texttt{provisioned}:

\begin{verbatim}
univorch /> status /lab/networks/student09
univorch /> # ... read the job history ...
univorch /> force-undeploy /lab/networks/student09
\end{verbatim}

The \texttt{force-undeploy} command is on the roadmap; until then,
a broken descriptor is cleared by hand at the hypervisor (delete
the residual VM if any) and a normal \texttt{undeploy} run from
the CLI.

\section{Monitoring and logs}

The container writes the daemon's standard output to Docker's
logging driver. Read the last hour with:

\begin{verbatim}
docker compose logs --since 1h -f
\end{verbatim}

For a long-running system, the Docker logs should be forwarded to
the institutional log collector (rsyslog, journald, ELK, Grafana
Loki) using the standard \texttt{logging:} stanza of
\texttt{docker-compose.yml}.

The orchestrator's \emph{operation} log is the persisted Job
history, not the standard output. The Job history is queried
through the CLI or the web GUI and survives restarts. The default
retention is 90~days; the value is configurable as a global
property.

\section{Backups and recovery}

The full state of UnivOrch is a single JSON file in the named
volume \texttt{univorch\_data} (the database is TinyDB in v1, a
choice that is replaced by MongoDB in production). A backup is a
copy of the volume.

The simplest periodic backup is a wrapper around \texttt{docker
cp}:

\begin{verbatim}
#!/bin/bash
set -e
ts=$(date +%F-%H%M)
docker cp univorch:/data/db.json /var/backups/univorch/univorch-$ts.json
\end{verbatim}

The recommended retention is GFS (DEC-024): seven daily backups,
four weekly backups, twelve monthly backups. A simple
\texttt{cron} entry plus a small rotation script implements it.

Restoring is also a file copy in v1. Stop the daemon, replace the
database file with the backup, start the daemon:

\begin{verbatim}
docker compose stop
docker cp /var/backups/univorch/univorch-2026-06-07-0300.json \
          univorch:/data/db.json
docker compose start
\end{verbatim}

When the persistence layer moves to MongoDB, the same workflow is
replaced by \texttt{mongodump} and \texttt{mongorestore} with
the same GFS schedule.

\section{Upgrades}

Upgrading the daemon to a new minor or patch release is a tag
change in \texttt{docker-compose.yml} followed by:

\begin{verbatim}
docker compose pull
docker compose up -d
\end{verbatim}

Docker downloads the new image, stops the running container and
starts a new one against the same persistent volume. The
operational data is preserved.

Major upgrades (\texttt{v1 → v2}) may include a migration step on
the database. The release notes of every major version describe
the migration; until that path is established, the upgrade is a
backup, a stop, an image swap, a migration command and a start.

\section{Users and authentication}

In v1 the daemon assumes the network it runs on is trusted: there
is no authentication on the REST API and the only role enforced in
code is the implicit \emph{superuser}. The full RBAC design lives
in \Cref{ch:conceptos} and is part of the immediate future work.

Until authentication is in place, the deployment should restrict
network exposure: a reverse proxy that only accepts requests from
the institutional VPN, a firewall rule that blocks the daemon's
port to the outside, or both.

\section{Troubleshooting}

\paragraph{The container does not start.}
\texttt{docker compose logs} usually shows the cause. The most
common ones are a busy port (change the host port mapping), a
permission problem on the volume (check the file owner) and a
missing image (the network blocks \texttt{ghcr.io}; use the
internal mirror).

\paragraph{The web GUI loads but the tree is empty.}
The database is empty. Load a definition with \texttt{load} as
described above.

\paragraph{Deploy fails immediately.}
The connector cannot reach the hypervisor. Verify the hypervisor's
address and credentials in the corresponding \texttt{define
hypervisors:} block, and check the network from inside the
container with \texttt{docker exec univorch curl -s
<hypervisor>}.

\paragraph{A descriptor sits in \texttt{unreachable}.}
The hypervisor was not reachable when an operation ran. The state
is reactive; once communications are restored, an operation on
that descriptor will move it back to a definite state. The Job
history records the original communication error.

\section{Where to find more information}

\begin{itemize}
  \item Source code: \texttt{https://github.com/clamaveruma/univorch}.
  \item Public image: \texttt{ghcr.io/clamaveruma/univorch}.
  \item REST API specification: \texttt{http://<server>:8080/api/v1/docs}
        (Swagger UI generated by FastAPI).
  \item Architecture and design: \Cref{ch:desarrollo} of this report
        and \texttt{docs/architecture.md} in the repository.
  \item Design-decision log: \Cref{ap:decisiones} and
        \texttt{claude/decisiones.md} in the repository.
\end{itemize}
