% Appendix C - Summary of technical decisions.
% Source: claude/decisiones.md (DEC-001 to DEC-036). The full text of
% each decision lives in the repository; here we only keep the summary
% the tribunal needs to follow the citations in the main body.

\umaappendix{Summary of technical decisions}
\label{ap:decisiones}

This appendix summarises the 36~technical decisions recorded during
the project. Each row lists the identifier, a one-line title, the
phase in which the decision was taken and the chapter of this report
where the decision is discussed in context. The full rationale of
every decision (motivations, alternatives considered, traceability
between decisions) lives in the project repository, in
\texttt{claude/decisiones.md}.

\section{Summary table}
\label{sec:dec-tabla}

\begin{longtable}{p{0.10\linewidth} p{0.62\linewidth} p{0.10\linewidth} p{0.10\linewidth}}
  \toprule
  \textbf{ID} & \textbf{Title} & \textbf{Phase} & \textbf{Chapter} \\
  \midrule
  \endfirsthead
  \toprule
  \textbf{ID} & \textbf{Title} & \textbf{Phase} & \textbf{Chapter} \\
  \midrule
  \endhead
  DEC-001  & Markdown for persistent memory                       & 1 & 3 \\
  DEC-002  & Context-directory structure (\texttt{claude/})       & 1 & 3 \\
  DEC-003  & Journal as primary source of truth                   & 1 & 3 \\
  DEC-004  & Two-layer architecture (engine + domain app)         & 1 & 2 \\
  DEC-005  & Terminology: ``descriptor''                          & 1 & 2 \\
  DEC-005b & Descriptor types (normal vs reference)               & 2 & 2 \\
  DEC-006  & Declarative architecture                             & 1 & 2 \\
  DEC-007  & Repository pattern for persistence                   & 3 & 5 \\
  DEC-008  & Use of design patterns (pragmatic)                   & 1 & 5 \\
  DEC-009  & End-user metaphor (\emph{desk} / \emph{computer})    & 1 & 2 \\
  DEC-010  & Cascade inheritance required from v1                 & 1 & 2 \\
  DEC-011  & RBAC with hierarchical inheritance, three roles      & 1 & 2 \\
  DEC-012  & Explicit imports + lexical closure of templates      & 2 & 2 \\
  DEC-013  & IP management out of v1                              & 1 & 5 \\
  DEC-014  & Operations model: Job pattern                        & 1 & 5 \\
  DEC-015  & Jobs persisted in the database                       & 1 & 5 \\
  DEC-016  & Hypervisor connector operations                      & 2 & 5 \\
  DEC-017  & Datastores as aliased resources                      & 1 & 5 \\
  DEC-018  & Client interfaces (CLI + Web GUI v1)                 & 1 & 5 \\
  DEC-019  & Project name (UnivOrch)                              & 1 & 1 \\
  DEC-020  & Access to pre-existing VMs (out of v1)               & 1 & 5 \\
  DEC-021  & User management (YAML in v1; LDAP future)            & 2 & 5 \\
  DEC-022  & Descriptor state machine (4 states + drift flag)     & 2 & 2 \\
  DEC-023  & Logs and operation retention                         & 2 & 5 \\
  DEC-024  & Database backup policy (GFS)                         & 2 & 5 \\
  DEC-025  & IP management: pools per folder                      & 2 & 5 \\
  DEC-026  & Inheritance rule by data type                        & 3 & 2 \\
  DEC-027  & Declarative load/plan with fail-fast validation      & 3 & 5 \\
  DEC-028  & Job engine: Command pattern, locking, sync v1        & 3 & 5 \\
  DEC-029  & Connectors as ABCs + registry                        & 3 & 5 \\
  DEC-030  & TinyDB $\rightarrow$ MongoDB; repositories per aggregate & 3 & 5 \\
  DEC-031  & Facade and centralised RBAC                          & 3 & 5 \\
  DEC-032  & Refined definition of \texttt{broken} state          & 3 & 5 \\
  DEC-033  & Technology stack v1                                  & 4 & 3 \\
  DEC-034  & Bounded adoption of Pydantic v2                      & 4 & 5 \\
  DEC-035  & Operation idempotency (in the orchestrator)          & 4 & 5 \\
  DEC-036  & Connector type registry + live-session pool          & 6 & 5 \\
  \bottomrule
\end{longtable}

\section{Structural decisions, in detail}

Six of the thirty-six decisions carry more weight than the rest
because they shape the architecture as a whole. They are presented
in full below; the references in parentheses point to the chapter
section where each is exercised.

\paragraph{DEC-004 — Two-layer architecture (\Cref{sec:dos-capas}).}
The system is split into a domain-agnostic engine (folders,
descriptors, jobs, inheritance, connectors) and a domain-specific
application that translates the vocabulary of a use case (teaching,
CTF, research) into operations of the engine. The boundary is the
public REST API of the daemon. The decision is the conceptual root
of the entire project: it is what lets the same engine support
multiple domain applications.

\paragraph{DEC-026 — Inheritance rule by data type (\Cref{sec:herencia}).}
Cascade inheritance applies a different merge rule depending on the
data type of the field. Scalars overwrite, lists accumulate,
dictionaries merge recursively. Specific fields that do not fit the
default (the IP pool, for example) declare an explicit override.
The rule is uniform across configuration properties \emph{and} role
assignments; the same code resolves both.

\paragraph{DEC-028 — Job engine: Command pattern, locking, synchronous v1.}
Every state-changing operation is encapsulated in a Command class.
The Command separates validation (returns the list of errors
without touching state) from execution (mutates state and may
raise). The engine wraps every Command in a Job that is persisted
before execution. The proof of concept executes Commands
synchronously; the architecture allows asynchronous execution with
locking without changes to the engine surface.

\paragraph{DEC-029 — Hypervisor connectors as ABCs (\Cref{subsec:patterns}).}
The contract between the engine and a hypervisor is defined as an
abstract base class with a small set of life-cycle operations
(\texttt{clone}, \texttt{delete}, \texttt{start}, \texttt{stop},
\texttt{get\_status} and a few more). The mock connector that ships
with v1 implements the same contract. A new hypervisor (VMware,
Proxmox) is added by writing a new concrete subclass; the engine
itself is not touched.

\paragraph{DEC-031 — Facade and centralised RBAC (\Cref{subsec:patterns}).}
The \texttt{OrchestratorService} class is the single entry point of
the engine and the only place where RBAC checks happen. Every
client (CLI, Web GUI, future teaching application) routes through
the facade; there is no shortcut.

\paragraph{DEC-036 — Connector type registry + live-session pool.}
The engine keeps two distinct structures. A registry of connector
\emph{types} maps a type name (\texttt{mock}, \texttt{vmware},
\texttt{proxmox}) to its connector class; this is what is patched
when a new hypervisor support is added. A pool of \emph{live
sessions} keeps the open connections to the configured hypervisors;
this is what survives a request and is reused across operations. The
decision was added late in the project after the architecture
diagrams revealed that the two concepts had been conflated in the
implementation.
